Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2015-12-16 jss/article.tex
[skip ci] jss update

Only the gene list is mandatory to call the \code{enrichment} function, all the
other arguments have default values.

[explain default values]\\
The \code{maps} argument can either be a dataframe imported from a gmt file with the \code{format_from_gmt} function or a list of dataframes generated by the same procedure. By default, the function uses the ACSN maps previously described.

The correction for multiple testing by default is set to false discovery rate (fdr), which is equivalent to Benjamini \& Hochberg correction, but can be changed to any of the usual corrections such as Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli \citep{Benjamini2003FDR}, or even disabled .

The minimal module size represents the smallest size value of a module that will be used to compute enrichment or depletion. This is meant to remove results of low significance for module of small size.

The universe in which the computation is made by default is defined by the maps. Which means that all genes that were given as input and that are not present on the maps will be discarded. To keep all genes, the user can set the universe to \code{HUGO},referring to the database.

The threshold is the maximal value of the corrected p-value (unless user chose not to correct for multiple testing) that will be displayed in the result table.

[give examples with maps, universe, correction multitest]

%multisample_enrichment<-function(Genes_by_sample=NULL,
%                                 maps = ACSNMineR::ACSN_maps, 
%                                 correction_multitest = "BH",
%                                 statistical_test = "fisher",
%                                 min_module_size = 5,
%                                 universe = "map_defined",
%                                 threshold = 0.05,
%                                 cohort_threshold = TRUE,
%                                 alternative = "greater"){
It may be of interest to compare enrichment of pathways in different cohorts or experiments. For example, enrichment of highly expressed pathways can reveal differences between two cancer types or two cell lines.
To facilitate such comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function. It relies on the \code{enrichment} function but takes a list of character vector genes. The name of each element of the list will be assumed to be the name of the sample for further analysis.
Most of the arguments given to \code{multisample_enrichment} are the same as the ones passed to \code{enrichment}. However, the \code{cohort_threshold} is designed to filter out modules which would not pass the significance threshold in all samples.

@@

In the simple example detailed below, we create a NaviCell session, then load
an expression data set from a local (tab-delimited) file. The data represent
gene expression measured in a prostate cancer cell line resistant to hormonal
treatment (agressive), and is taken from the Cell Line Encyclopedia project
\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
(the default map), using heat maps. With this visualization mode, gene
expression values are represented as a color gradient (green to red) in
squares positioned next to the entities where the gene has been mapped.

\begin{verbatim}
# load RNaviCell library

@@

@@

@@

\begin{figure}[!ht]
  \caption{Gene expression values from a prostate cancer cell line visualized
on the cell cycle map as heat map plots.}
  \centering
    \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
\end{figure}

>>>>

Only the gene list is mandatory to call the \code{enrichment} function, all the
other arguments have default values.  The \code{maps} argument can either be a
dataframe imported from a gmt file with the \code{format_from_gmt} function or
a list of dataframes generated by the same procedure. By default, the function
uses the ACSN maps previously described.  The correction for multiple testing
is set by default to false discovery rate (FDR), which is equivalent to
Benjamini \& Hochberg correction, but can be changed to any of the usual
correction methods (Bonferroni, Holm, Hochberg, Holm, or Benjamini \& Yekutieli
\citep{Benjamini2003FDR}), or even disabled .  The minimal module size
represents the smallest size value of a module that will be used to compute
enrichment or depletion. This is meant to remove results of low significance
for module of small size.  The universe in which the computation is made by
default is defined by all the gene symbols conatined in the maps. All the genes
that were given as input and that are not present on the maps will be
discarded. To keep all genes, the user can change the universe to \code{HUGO},
and in that case, the complete list of HUGO gene symbols will be used as the
reference ($>$ 39,000 genes). The threshold corresponds to the maximal value of the
corrected p-value (unless the user chose not to correct for multiple testing) that
will be displayed in the result table.

%multisample_enrichment<-function(Genes_by_sample=NULL,
%                                 maps = ACSNMineR::ACSN_maps, 
%                                 correction_multitest = "BH",
%                                 statistical_test = "fisher",
%                                 min_module_size = 5,
%                                 universe = "map_defined",
%                                 threshold = 0.05,
%                                 cohort_threshold = TRUE,
%                                 alternative = "greater"){

It may be of interest to compare enrichment of pathways in different cohorts or
experiments. For example, enrichment of highly expressed pathways can reveal
differences between two cancer types or two cell lines.  To facilitate such
comparisons, \pkg{ACSNMineR} provides a \code{multisample_enrichment} function.
It relies on the \code{enrichment} function but takes a list of character
vector genes. The name of each element of the list will be assumed to be the
name of the sample for further analysis.  Most of the arguments given to
\code{multisample_enrichment} are the same as the ones passed to
\code{enrichment}. However, the \code{cohort_threshold} is designed to filter
out modules which would not pass the significance threshold in all samples.

@@

In the simple example detailed below, we create a NaviCell session, then load
an expression data set from a local (tab-delimited) file. The data represent
gene expression measured in a prostate cancer cell line resistant to hormonal
treatment (agressive), and is taken from the Cell Line Encyclopedia project
\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
(the default map), using heat maps. With this visualization mode, gene
expression values are represented as a color gradient (green to red) in
squares positioned next to the entities where the gene has been mapped (Figure~\ref{fig:du145}).

\begin{figure}[!ht]
  \caption{Gene expression values from a prostate cancer cell line visualized
on the cell cycle map as heat map plots. The figure is a screenshot of the
NaviCell map browser, with the map set at the top (the less detailed) zoom
level. The essential phases of the cell cycle are indicated on the map
(G1/S/G2/M).} 
  \centering
  \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
  \label{fig:du145}
\end{figure}

\begin{verbatim}
# a short RNaviCell script example

# load RNaviCell library

@@

# import a gene expression matrix and 
# send the data to the NaviCell server
# NB: the data_matrix object is a regular R matrix

@@

# set data set and sample for heat map representation

@@
